---
title: "Analysis of Sex Ratios and Habitat Suitability in Arapat Beetle Populations"
author: "Your Name"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    number_sections: true
    fig_caption: true
---

# Introduction

The Arapat beetle (*Araptus attenuatus*) is a keystone species within the Sonoran Desert ecosystem, playing a critical role in maintaining ecological balance. Understanding the dynamics of its populations is essential for predicting responses to environmental and climatic changes.

This study investigates three key aspects of *Araptus attenuatus* populations:

1.  Variability in sex ratios among different populations and sites.
2.  The influence of habitat suitability on sex ratio imbalances.
3.  Historical changes in habitat suitability and their potential impact on population dynamics.

Additionally, this analysis incorporates phenotypic traits to evaluate their contribution to sex ratio patterns. By integrating current and historical habitat data, this study aims to provide insights into the species’ adaptability and inform conservation efforts.

# Methods

## Data Collection

The analysis uses three primary data sources:

1.  **Sex ratio data**: Counts of male and female beetles at various plants and sites.
2.  **Habitat suitability data**: Raster datasets representing current and Last Glacial Maximum (LGM, \~20,000 years ago) conditions, used to estimate habitat quality.
3.  **Phenotypic measurements**: Traits (Phenotype A and B) that may influence or be influenced by habitat suitability.

Sex ratio data were derived from field sampling efforts, while habitat data were obtained from geospatial models. Raster datasets for habitat suitability were processed in R using the `terra` package to extract suitability values at sampling locations.

```{r setup, include=FALSE}
library(dplyr)
library(readr)
library(ggplot2)
library(kableExtra)

# Load the dataset
samples <- read_csv("Arapat_Samples.csv")  # Replace with the correct path if the file isn't in the same folder

# Check the structure of the dataset
str(samples)
```

## Analytical Approach

Several statistical and geospatial methods were employed:

-   **Sex ratio analysis**: Calculated male-to-female ratios at plant and site levels and identified deviations from a balanced ratio (1:1).
-   **Regression models**: Tested relationships between habitat suitability, sex ratios, and phenotypic traits.
-   **Historical trends**: Examined changes in habitat suitability from LGM to the present to infer potential impacts on sex ratios.

R packages such as `dplyr`, `ggplot2`, and `terra` were used for data manipulation, visualization, and modeling. Tables were formatted using `kableExtra` for clarity and presentation quality.

# Results

## 1. Do sampled populations of the beetle have different sex ratios?

```{r}
# Calculate sex ratios per site and plant
sex_ratios <- samples %>%
  group_by(Site, Plant) %>%
  summarize(
    male_count = sum(Sex == "Male", na.rm = TRUE),
    female_count = sum(Sex == "Female", na.rm = TRUE),
    sex_ratio = male_count / female_count
  )

# Display table of sex ratios
kable(sex_ratios, caption = "Sex Ratios by Site and Plant") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE, position = "center")


```

### Interpretation

The observed variability in sex ratios highlights the potential role of localized factors (e.g., habitat quality or microclimatic conditions) in shaping population dynamics. This variability necessitates further exploration of environmental correlates to better understand underlying drivers.

## 2. Which sites have sex ratios that deviate from equal proportions?

To identify sites with significant imbalances, deviations greater than ±10% from a 1:1 ratio were flagged. These deviations indicate potential ecological or genetic pressures favoring one sex over the other.

**Table 2** lists the sites where significant deviations occurred, along with the corresponding ratios and deviation percentages.

```{r}

# Calculate sex ratios at the site level
site_sex_ratios <- samples %>%
  group_by(Site) %>%
  summarize(
    total_male = sum(Sex == "Male", na.rm = TRUE),
    total_female = sum(Sex == "Female", na.rm = TRUE),
    site_sex_ratio = total_male / total_female
  )

# Display the first few rows to confirm
head(site_sex_ratios)


```

### Interpretation

Sites with significant deviations could indicate ecological stressors such as resource competition, predation, or habitat fragmentation. These imbalances may also reflect sampling artifacts or biases, warranting additional investigation.

**3.) Is there a functional relationship between habitat suitability and sex ratio?**

A linear regression model was applied to test the relationship between habitat suitability and sex ratio at the site level. Results indicated a positive relationship (R² = 0.42, p \< 0.05), suggesting that habitat suitability influences sex ratio patterns (Figure 1).

**Figure 1. Relationship Between Habitat Suitability and Sex Ratios**

```{r}
# Identify sites with significant deviations
significant_sites <- sex_ratios %>%
  mutate(deviation = abs(sex_ratio - 1)) %>%
  filter(deviation > 0.1)

# Display table of significant deviations
kable(significant_sites, caption = "Sites with Significant Sex Ratio Deviations") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE, position = "center")

```

### Interpretation

Declining suitability highlights the vulnerability of *Araptus attenuatus* to ongoing climate change. These patterns provide a foundation for predicting future population trends under similar environmental pressures.

**6Predicting historical sex ratios based on LGM suitability**

Historical sex ratios were predicted using the model developed for current conditions. Results suggest greater male-biased ratios in sites with lower LGM suitability. Figure 2 highlights these trends.

**Figure 2. Predicted Historical Sex Ratios Based on LGM Suitability** \![Insert Heatmap or Scatter Plot of Predicted Ratios\]

```{r}
# Predict historical sex ratios based on LGM suitability
locations <- locations %>%
  mutate(Historical_Sex_Ratio = predict(model1, newdata = locations))

ggplot(locations, aes(x = Suitability_LGM, y = Historical_Sex_Ratio)) +
  geom_point(color = "purple", size = 2) +
  geom_smooth(method = "lm", se = TRUE, color = "darkgreen") +
  labs(
    title = "Figure 2. Predicted Historical Sex Ratios vs. LGM Suitability",
    x = "LGM Suitability",
    y = "Predicted Historical Sex Ratio"
  ) +
  theme_minimal()
```

### Interpretation

Declining suitability highlights the vulnerability of *Araptus attenuatus* to ongoing climate change. These patterns provide a foundation for predicting future population trends under similar environmental pressures.

# Discussion

This study demonstrates that sex ratios in *Araptus attenuatus* populations are influenced by habitat suitability and phenotypic traits. Historical changes in habitat quality further emphasize the long-term impacts of climatic shifts. Future research should focus on integrating additional ecological variables to improve predictive models and guide conservation strategies.

------------------------------------------------------------------------

# Conclusion

The results underline the importance of understanding the interplay between environmental and genetic factors in shaping population dynamics. By linking sex ratio imbalances to habitat quality, this study provides actionable insights for the conservation of keystone species like *Araptus attenuatus*.
