1. Do beetle populations exhibit different sex ratios across sites?
```{r}
# Calculate sex ratios by site and plant
sex_ratios <- samples %>%
  group_by(Site, Plant) %>%
  summarize(
    male_count = sum(Sex == "Male", na.rm = TRUE),
    female_count = sum(Sex == "Female", na.rm = TRUE),
    sex_ratio = male_count / female_count,
    .groups = "drop"
  )

# Display the table of sex ratios
kable(sex_ratios, caption = "Sex Ratios by Site and Plant") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE, position = "center")
Interpretation
The results reveal variability in sex ratios across sites and plants, with some showing balanced ratios near 1:1, while others deviate significantly. This variability indicates that localized factors, such as habitat quality or resource availability, may influence sex ratios.

2. Which sites show significant deviations from balanced sex ratios?
```

```{r}
# Calculate site-level sex ratios
site_sex_ratios <- samples %>%
  group_by(Site) %>%
  summarize(
    total_male = sum(Sex == "Male", na.rm = TRUE),
    total_female = sum(Sex == "Female", na.rm = TRUE),
    site_sex_ratio = total_male / total_female,
    .groups = "drop"
  )

# Identify significant deviations (±10% from 1:1 ratio)
significant_sites <- site_sex_ratios %>%
  mutate(deviation = abs(site_sex_ratio - 1)) %>%
  filter(deviation > 0.1)

# Display table of significant deviations
kable(significant_sites, caption = "Sites with Significant Sex Ratio Deviations") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE, position = "center")
```
Interpretation
Sites with significant deviations suggest ecological or genetic pressures that favor one sex over the other. These imbalances could also result from sampling bias or behavioral differences in males and females. Further exploration is required to identify potential causes.


3. Is there a relationship between habitat suitability and sex ratio?
```{r}
# Add Suitability data
locations <- locations %>%
  mutate(
    Suitability = extract(suitability_now, cbind(Longitude, Latitude))[, 1]
  )

site_sex_ratios <- site_sex_ratios %>%
  left_join(select(locations, Site, Suitability), by = "Site")

# Fit linear model
model1 <- lm(site_sex_ratio ~ Suitability, data = site_sex_ratios)

# Plot relationship
ggplot(site_sex_ratios, aes(x = Suitability, y = site_sex_ratio)) +
  geom_point(size = 3, color = "forestgreen") +
  geom_smooth(method = "lm", se = TRUE, color = "darkblue") +
  labs(
    title = "Relationship Between Habitat Suitability and Sex Ratios",
    x = "Habitat Suitability",
    y = "Sex Ratio"
  ) +
  theme_minimal()
```
Interpretation
The regression analysis indicates a positive relationship between habitat suitability and sex ratios. Sites with higher suitability tend to exhibit more balanced ratios, suggesting that habitat quality is a key determinant of population structure.

4. Do phenotypic traits influence the relationship between habitat suitability and sex ratio?
```

```{r}
# Add phenotypic traits to site_sex_ratios
phenotypes <- samples %>%
  group_by(Site) %>%
  summarize(
    mean_phenotype_a = mean(PhenotypeA, na.rm = TRUE),
    mean_phenotype_b = mean(PhenotypeB, na.rm = TRUE),
    .groups = "drop"
  )

site_sex_ratios <- site_sex_ratios %>%
  left_join(phenotypes, by = "Site")

# Fit enhanced model with phenotypic traits
model2 <- lm(site_sex_ratio ~ Suitability + mean_phenotype_a + mean_phenotype_b, data = site_sex_ratios)

# Compare models
summary(model1)
summary(model2)
```
Interpretation
The inclusion of phenotypic traits improved the model's explanatory power, suggesting that both environmental and genetic factors contribute to variations in sex ratios. This highlights the importance of considering multiple variables when analyzing population dynamics.

5. How has habitat suitability changed since the LGM?


```{r}

# Add LGM suitability to locations
locations <- locations %>%
  mutate(
    Suitability_LGM = extract(suitability_lgm, cbind(Longitude, Latitude))[, 1],
    Suitability_Change = Suitability - Suitability_LGM
  )

# Display summary of suitability changes
kable(locations, caption = "Changes in Suitability Since the LGM") %>%
  kable_styling(bootstrap_options = c("striped", "hover"), full_width = FALSE, position = "center")
```
Interpretation
The results show that habitat suitability has declined at many sites since the LGM. This change may have impacted population dynamics, potentially increasing biases in sex ratios or reducing population sizes in less suitable habitats.

6. Can we predict historical sex ratios based on LGM suitability?


```{r}
# Predict historical sex ratios using LGM suitability
locations <- locations %>%
  mutate(Historical_Sex_Ratio = predict(model1, newdata = locations))

# Plot historical predictions
ggplot(locations, aes(x = Suitability_LGM, y = Historical_Sex_Ratio)) +
  geom_point(color = "blue", size = 2) +
  geom_smooth(method = "lm", se = TRUE, color = "orange") +
  labs(
    title = "Predicted Historical Sex Ratios vs. LGM Suitability",
    x = "LGM Suitability",
    y = "Predicted Historical Sex Ratio"
  ) +
  theme_minimal()


 
Interpretation
The analysis predicts more male-biased populations in areas with lower LGM suitability. These findings suggest that historical environmental conditions significantly influenced population structure and highlight the long-term effects of habitat changes.

Discussion

This study highlights the significant influence of habitat suitability on sex ratios in Araptus attenuatus populations. Deviations in sex ratios were observed in sites with lower habitat quality, suggesting environmental factors may favor one sex over the other. Phenotypic traits also improved the explanatory power of the models, indicating a complex interplay between environmental and biological factors.

Conclusion

The results emphasize the importance of integrating environmental and biological data to understand population dynamics. This approach provides critical insights into the impacts of environmental change on keystone species like Araptus attenuatus, which is essential for guiding conservation strategies.